Linear stability of magnetized massive protoplanetary disks

### 2014-03-27 paper/paper.tex
tex update
tex updates

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
\newcommand{\zhat}{\hat{z}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\zmax}{Z_s}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
%\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\dphi}{\dd \Phi}
\defcitealias{lin12}{L12}

@@

\author{Min-Kai Lin (+Menou, Fromang, Youdin)}
\affil{Canadian Institute for Theoretical Astrophysics,
  60 St. George Street, Toronto, ON, M5S 3H8, Canada}
\email{mklin924@cita.utoronto.ca}

\begin{abstract}

\end{abstract}

\section{Introduction}
Understanding the evolution of protoplanetatry disks (PPDs) is key to planet
formation theory. A central aspect of PPD evolution is the mechanism
of angular momentum transport.

\input{setup}
\input{linear}
\input{results}
\input{results_gi}
\input{results_azi}

>>>>

\usepackage{amsmath}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
\newcommand{\zhat}{\hat{z}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\zmax}{Z_s}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\w}{ \widetilde{W}}
%\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12}

@@

\author{Min-Kai Lin }%(+Menou, Fromang, Youdin)}
\affil{Canadian Institute for Theoretical Astrophysics,
  60 St. George Street, Toronto, ON, M5S 3H8, Canada}
\email{mklin924@cita.utoronto.ca}

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop
  both. The goal of this paper is to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The linear stability of a local patch of a
  self-gravitating, magnetized three-dimensional (3D) disk to
  axisymmetric perturbations is examined. %Addition effects such as
  %resistivity and azimuthal fields are considered.

%We examine the local linear stability of
%  magnetized, self-gravitating and differentially rotating disks to
%  axisymmetric perturbations. 
%
%not many studies model both
%revisits 
\end{abstract}

\section{Introduction}
Astrophysical disks host a wide range of fluid instabilities. Among
them, the magneto-rotational instability \citep[MRI,][]{chandrasekhar61,
  balbus91,balbus98} %balbus %chandrasekhar ref
and gravitational instability \citep[GI, ][]{toomre64,goldreich65a,goldreich65b} 
provide robust pathways to turbulent angular momentum transport that
enables mass accretion \citep[][ 
  and references therein]{balbus99,armitage11,turner14}.   
%gammie, lin/pringle, paczynski 
%relevance of MRI and GI to planet formation 
They are also relevant to planet formation theory. For example,
the strength of MRI turbulence directly affect planetesimal dynamics
in protoplanetary disks \citep{yang12,gressel12}; while GI can
potentially form giant planets directly through disk fragmentation
\citep{boss97,boss98,gammie01,voro13,helled14}. 
%gammie,
%boss %some review articles (PP6)  
%shakura73, pacz78, lin87,

Accretion disks such as those surrounding black holes can develop both
MRI and GI \citep{menou01,goodman03}.  
Protoplanetary disks (PPDs) are also expected to be massive and  
magnetized in its earliest evolutionary phase \citep{inutsuka10}.  
The interplay between MRI and GI has been invoked to explain outbursts
in circumstellar disks \citep{armitage01, zhu10a, zhu10b,
  martin12b}, and predicts similar phenomenon in circumplanetary disks
\citep{lubow12}. This results from the development of `dead zones' 
--- magnetically inactive, laminar regions near the disk midplane --- with
magnetized layers above and below 
\citep{gammie96,martin12,landry13}. Mass accumulation in the dead zone can lead to
GI and trigger MRI through heating. In these models, the condition required
for MRI is realized through GI, but the MRI is unaffected by disk
self-gravity.

PPDs subject to both MRI and GI are often modeled through separate
turbulent viscosities in a hydrodynamical framework 
\citep{terquem08}. This implicitly assumes that the development of MRI 
and GI can be assessed independently. 
%That is, MRI is unaffected by
%disk self-gravity; and GI is unaffected by weak magnetization. 
Circumstellar disk models that explicitly combine the equations of magneto-hydrodynamics
and self-gravity have been limited to a few early simulations 
\citep{fromang04,fromang04b,fromang05}. It will be neccessary to %no
%major difference % kim/ostriker/stone
revisit and extend these pioneering calculations to fully explore the
impact of MRI and GI on the structure and evolution of PPDs. In
preparation of this, it is important to have a thorough understanding
of the stability properties of such systems.  
%non-ideal effects important

%why is GI and MRI independent 
%compressibility is not fundamental to MRI 
%horizontal wavenumber (gammie/balbus) 
%The effect of self-gravity on the linear MRI has not been considered, 
%since fluid compressibility is not fundamental to the MRI.

Since compressibility is not fundamental for the MRI, much of the
early stability calculations assume incompressible perturbations
\citep{goodman94,jin96}. However, recent works indicate
compressibility may be important under certain conditions,
such as strong fields \citep{kim00, pessah05,bonanno07}.   
Previous MRI studies have also focused on modes with vanishing
radial wavenumber, because they are the most unstable
\citep{sano99,reyes01}. Self-gravity has minimal effect on such
perturbations in a rotating disk. However, modes
with radial lengthscales on the order of the disk scale height may be
subject to self-gravity. It is of interest to generalize the MRI with
non-zero radial waveumbers to massive disks.

%Simiarly, studies of GI in accretion disks have mostly neglected
%magnetization \citep{mamat10,kim12}. 
%particularly in the context of

%Studies of GI in accretion disks do not often include magnetization
%\citep{mamat10,kim12}.

The effect of a magnetic field on the GI of rotating disks has been
considered by some authors.  
% \cite{lynden-bell66} found that a toroidal
%field can destabilize non-axisymmetric disturbances
Recently, \cite{lizano10} generalized the classic Toomre stability
criterion for razor-thin disks to include a vertical field. For
circumstellar disks, the authors concluded that the field is
stabilizing. This is consistent with previous analysis by
\cite{nakamura83} for three-dimensional (3D) uniformly rotating
disks. However, the GI of 3D differentially rotating disks have
mostly neglected magnetic fields \citep{mamat10,kim12}, but such disks
are subject to the MRI if magnetized. 
%how is the dispersion relation for GI affected by the presence of MRI?
%It is interesting to ask whether
%or not the presence of MRI affects GI. 
%
%It is desirable to clarify
%whether or not the presence of MRI modes affect the dispersion
%relation for GI.

%although MRI and GI are physically distinct 
%explicitly show

This paper marks the beginning of our study of magnetized,
self-gravitating PPDs. We start from linear calculations. 
Linear stability analyses have the advantage that a wide range of
parameters can be studied with negligible computational cost. This
allows one to indentify conditions, if any, under which MRI and GI
cannot be consdered independent. It is also important to have such 
calculations at hand to help benchmark and guide non-linear
simulations.

%This paper marks the beginning of our study of magnetized,
%self-gravitating PPDs and is organized as follows.

This paper is organized as follows. \S\ref{setup} lists the governing
equations and describes the disk equilibra under consideration. 
The linear problem is formulated in \S\ref{linear}. The impact
of self-gravity on the MRI is discussed in
\S\ref{result1}; gravitationally unstable disks are considered in
\S\ref{result2} and equilibra with an azimuthal field is explored in
\S\ref{result3}. We summarize results in \S\ref{summary}, including a
discussion of caveats and future directions.

\input{setup}
\input{linear}
\input{results}
\input{results_gi}
\input{results_azi} 
\input{summary}

### 2014-03-27 paper/results_azi.tex
tex updates

\section{Effect of an azimuthal field}
We now extend the previous investigation to disks with an azimuthal
field, so that $B_y\neq 0$.

%cannot consider superthermal fields
%compressibility important f

%\subsection{}

>>>>

\section{Effect of an azimuthal field}\label{result3}
In this section we add an azimuthal field so that $B_y\neq 0$,  
parameterized by $\epsilon \equiv B_y/B_z$. However, we continute to
use $B_z$ for normalizations and $\beta$ is associated with the
vertical Alfven speed. We also extend the previous calculations to the
full disk $z\in[-Z_s,Z_s]$, which allows us to compare the effect of
self-gravity on MRI modes with different symmetries across the
midplane.

\subsection{Ideal disks without GI} 
Here we consider an isothermal disk with $Q=0.2$
($Q_\mathrm{2D}=0.72$) and $\beta=10$ in the limit of ideal MHD
($\Lambda_0=100$). Gravitational instability is not expected because 
Fig. \ref{compare_growth3} shows that even for $Q=0.18$, GI is 
surpressed for $\beta \lesssim 15$. We consider azimuthal field
strengths with $\epsilon=[0,3]$.

Fig. shows the MRI growth rates for the above system as a function of
$k_x$. The top panel are modes with even midplane symmetry: 
$W^\prime(0)\sim 0$ for $k_x\to0$, and the bottom panel are modes with odd midplane 
symmetry: $W(0)\sim 0$ for $k_x\to0$. The latter set of modes were 
excluded in the previous sections by midplane boundary 
conditions. The plot also include growth rates computed in the Cowling
approximation. As expected, none of the modes are energetically
dominated by self-gravity, $\avg{E_g}<\avg{E_m}$.

%general stabilization of MRI

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm 0cm]{figures/compare_growth3_tilted_even.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0.52cm]{figures/compare_growth3_tilted_odd.ps}
  \caption{Growth rates in massive isothermal disks with $Q=~0.2$ and
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. The
    top panel correspond to even modes and the
    bottom panel correspond to odd modes, which have
    $W^\prime(0)\sim0$ and $W(0)\sim0$, respectively, for $k_x\to0$.
    \label{compare_growth3_tilted}}
\end{figure}

%We consider $\epsilon\equiv B_y/B_z \in[0,3]$. The

Consider first the even modes in the top panel of
Fig. \ref{compare_growth3_tilted}. Self-gravity slightly stabilizes
modes with small $k_x$ as $B_y$ is increased. SG destabilizes modes
with  $k_xH\gtrsim O(1)$, and is most effective for purely
vertical fields: for $\epsilon=0,\, k_xH\simeq 1.4$, SG increases the
growth rate by $\sim 30\%$. Consequently the cut-off wavenumber is
larger when SG is included. This destabilization effect weakens with
$B_y$ because the total magnetic pressure increases, which generally
opposes GI, although the modes are fundamentally associated with MRI.

The odd modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. Self-gravity has negligible effect when 
the field is purely vertical ($B_y=0$). SG stabilizes all wavelengths,
and is most effective with increasing $B_y$ and $k_xH =  O(1)$.

%With $B_y\neq 0$, the MRI is compressible even in the Cowling
%approximation \citep{pessah05}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0cm]{figures/result_tilted_cowling.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/result_tilted_fullsg.ps} 
  \caption{caption
    \label{result_tilted}}
\end{figure}

\subsection{Resistive disks with GI}

%\subsection{GI}

### 2014-03-27 paper/results_gi.tex
tex update
tex updates

\section{Gravitationally unstable disks} 
Gravitational instability (GI) becomes possible in a sufficiently
massive and/or cold disk. We explore how the emergence of GI affects
the MRI, and vice versa. To do so, we compute unstbale modes for  
isothermal disks with $Q < 0.2 $ ($Q_\mathrm{2D}\lesssim 0.67$) which
permits gravitational instability as well as the MRI.

\subsection{Ideal disks}
We first consider ideal MHD by setting $\Lambda_0=100$ and $A=1$. 
Fig. \ref{compare_growth3} show growth rates for 
modes with $k_xH=1$ as a function of $\beta$ in disks with $Q=0.18$,
$Q=0.14$ and $Q=0.12$. The colorbar quantifies the importance of
self-gravity with 
\begin{align}
 \tau \equiv \frac{\avg{E_g}}{\avg{E_g} + \avg{E_m}}. 
\end{align} 
Thus, modes with $\tau  = 1$ are
energetically dominated by self-gravity (GI) and $\tau\ll1$ represents a mode
dominated by magnetic perturbations (MRI).

@@

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=0.18$
where the GI branch approaches an MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. 
%Although the MRI is still most unstable, it is
%self-gravitating for $\beta\lesssim 50$ (uppermost curve). 
Some modes may have been missed in our
numerical search of eigenfrequencies. Thus, `breaks' along the
GI branch for $Q=0.18$ and $Q=0.12$ does not prove that the GI and MRI branches do not
intersect. However, the continous variation of $\tau$ with respect to
$\beta$ suggests that unstable modes can transition smoothly from
magnetically-dominated to gravitationally-dominated at low $\beta$.

For $Q=0.12$ the GI growth rate exceeds all MRI modes. Notice the most
unstable MRI branch becomes self-gravitating for $20 \lesssim
\beta\lesssim 40$, but magnetic perturbations remain negligible
for GI. 
%we generally find MRI modes aquire significant potential
%perturbations but GI modes always have negligible magnetic
%perturbation. 
Fig. \ref{compare_growth3_Q01d2} show growth rates
as a function of $k_xH$ for the $Q=0.12,\,\beta=20$ disk. All
perturbations with $k_xH \lesssim 3.5$ are unstable. MRI operates on the largest
length scale, followed by GI at $0.7 \lesssim k_xH 1.7$. The 
transition between MRI to GI is smoother with increasing $k_xH$ as the
MRI becomes compressible.

%Modes with $k_xH
% \gtrsim 1.7$ are neither pure MRI or GI. 
%GI wave lengths fit inside channel solutions

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity. The solid line corresponds to MRI modes in the
    Cowling approximation. 
    \label{compare_growth3_Q01d2}}
\end{figure}

\subsection{Layered disks}
Can a disk with a dense, resistive midplane and a highly conducting
atmosphere  be subject to both MRI and GI but at different heights?

>>>>

\section{Gravitationally unstable disks} \label{result2}
Gravitational instability (GI) becomes possible in a sufficiently
massive and/or cold disk. We explore how the emergence of GI affects
the MRI, and vice versa. To do so, we compute unstbale modes for  
isothermal disks with $Q < 0.2 $ ($Q_\mathrm{2D}\lesssim 0.67$) which
permits gravitational instability as well as the MRI.

\subsection{Co-existence of MRI and GI}
We first consider ideal MHD by setting $\Lambda_0=100$ and $A=1$. 
Fig. \ref{compare_growth3} show growth rates for 
modes with $k_xH=1$ as a function of $\beta$ in disks with $Q=0.18$,
$Q=0.14$ and $Q=0.12$. The colorbar quantifies the importance of
self-gravity with 
\begin{align}
 \tau \equiv \frac{\avg{E_g}}{\avg{E_g} + \avg{E_m}}. 
\end{align} 
Thus, modes with $\tau  = 1$ are
energetically dominated by self-gravity (GI) and $\tau\ll1$ represents a mode
dominated by magnetic perturbations (MRI).

@@

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=0.18$
where the GI branch approaches an MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to avoided crossing of modes, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori 
\citep[e.g.][]{christo93}. However, we cannot exclude the possibility 
that some modes may have been missed in the numerical search of
eigenfrequencies.

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continous variation of $\tau$ with
respect to $\beta$ strongly suggest that unstable modes can transition
smoothly from magnetically-dominated to gravitationally-dominated at
low $\beta$.

\subsubsection{Case study} 
In reality, perturbations with a range of $k_x$ will be present for a
given set of disk parameters. Fig. \ref{compare_growth3_Q01d2} show
growth rates as a function of $k_x$ for an isothermal disk with
$Q=0.12,\,\beta=20$. MRI and GI are active with comparable growth
rates. All perturbations with $k_xH \lesssim 3.5$ grow dynamically
($\gamma\gtrsim 0.1\Omega$, or $\lesssim$ 1.6 orbits).

We also plot in Fig. \ref{compare_growth3_Q01d2} growth rates obtained 
from the Cowling approximation, which isolates MRI; and that from a
high-resistivity run, which isolates GI by allowing the field lines to
slip through the fluid. We refer to these as  
pure MRI and pure GI, respectively. 
For $k_xH\lesssim 0.2$, growth rates are equal to
those on the pure MRI and pure GI branches. That is, MRI and GI
operate independently until their growth rates become equal as a
function of $k_x$.

The dispersion relation $\gamma(k_x)$ increasingly deviates from the
pure GI/MRI curves with $k_x$, implying increasing interaction between
magnetic and density perturbations. 
Comparing pure GI (dashed line)  and the gravitationally-dominated
portions of $\gamma(k_x)$ indicate inclusion of magnetic field
stabilizes high-$k_x$ pure GI. (Note also the slight decrease in the most unstable
$k_x$.) Since axisymmetric GI is unaffected by shear, this
stabilization is due to magnetic pressure and is consistent with
\citep{lizano10}.

%3D geometry doesn't affect 
%axisymmetric GI doesn't care about shear 
%
%two legs of GI are not symmetric 
%MRI become self-gravitating
%GI with increasing magnetization

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity can enhance or reduce
MRI growth rates depending on $k_x$, but the effect is small.

%magnetic pressure, small scale more effective 
%, reflected in the full range of $\tau$ displayed by the
%modes. 
%mode classification becomes difficult. 
%the point is: dispersion relation cannot be estimated by computing
%the MRI and GI separately

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resisitivity in the full problem. 
    \label{compare_growth3_Q01d2}}
\end{figure}

%\subsection{Layered disks}
%Can a disk with a dense, resistive midplane and a highly conducting
%atmosphere  be subject to both MRI and GI but at different heights?
